The unequal distribution of access to education and transport: insights from

### 2016-05-31 paper.Rmd
Reword introduction

Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas. In this paper, we propose a new school accessibility index that combines information of the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure. We calculate a school accessibility index for 633 areas within the SPMR and use it to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility. We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities. (INCLUDE SOMETHING ABOUT CONTESTATION)

The new index we propose here constitutes a methodological contribution to the measurement of  school accessibility. It is inspired by the index developed by Shen (1998) in the context of job accessibility. The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area. Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available. As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas. The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas. The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unequally across space.

>>>>

Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas. In this dynamic and highly political context, new quantitative evidence can help shed light on the the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing. Moreover, the results should help design policies that are more inclusive and equitable by highlighting areas where the provision of schools, travel to school opportunities, or both, are particularly restrictive.

Motivated by this political context this paper presents a new school accessibility index. It combines information about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure. We calculate a school accessibility index for 633 areas within the SPMR and use it to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility. We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.

<!-- (INCLUDE SOMETHING ABOUT CONTESTATION) -->

The new index contributes to the measurement of school accessibility. It is inspired by the index developed by Shen (1998) in the context of job accessibility. The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area. Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available. As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas. The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas. The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unequally across space.
